\documentclass[a4paper,twoside]{article}

% fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage[scaled=0.8]{beramono}

% geometry
\usepackage[a4paper,centering]{geometry}
\setlength{\marginparwidth}{2cm}

% Inria research report
\usepackage{RR}

% extra packages

\usepackage{hyperref}
\usepackage{xspace}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}

\usepackage[color=yellow!45!white,linecolor=black,textsize=footnotesize]{todonotes}

\usepackage{why3lang}

% placement of figures
\renewcommand\floatpagefraction{0.99}
\renewcommand{\textfraction}{0.01}
\renewcommand{\topfraction}{0.99}
\renewcommand{\bottomfraction}{0.99}
\setcounter{topnumber}{4}
% \setcounter{bottomnumber}{4}
% \setcounter{totalnumber}{8}


\newcommand{\ie}{\textit{i.e.\/}\xspace}
\newcommand{\eg}{\textit{e.g.\/}\xspace}


\RRNo{XXXX}    % TODO
% \RTNo{XXXX}
%%
%% date de publication du rapport
\RRdate{xxx 2024}
%%
%% Cas d'une version deux
%% \RRversion{2}
%% date de publication de la version 2
%% \RRdater{November 2008}
%%
\RRauthor{% les auteurs
  Claude Marché\thanks[toccata]{Université Paris-Saclay, CNRS, ENS Paris-Saclay, Inria,
    LMF, 91190, Gif-sur-Yvette, France}
}
%% Ceci apparait sur chaque page paire.
\authorhead{Marché}
%% titre francais long
\RRtitle{Un état de l'art des modèles mémoires pour les programmes à pointeurs\thanks{Ces recherches ont été partiellement financées par Décysif TODO....}}
%% English title
\RRetitle{A state of the art of memory models for pointer programs\thanks{This work has been partially supported by Décysif... TODO}}
%%
\titlehead{Memory Models for Pointer Programs}
%%
% \RRnote{This is a note}
% \RRnote{This is a second note}
%%
\RRresume{A FAIRE}

\RRabstract{TO DO}

%%
\RRmotcle{Vérification déductive, Programmes à pointeurs, Modèles mémoire}
%
\RRkeyword{Deductive Verification, Pointer programs, Memory models}
%%
\RRprojet{Toccata}  % cas d'un seul projet
% \RRprojets{Apics and Op\'era and Marelle}
%%
%% \URLorraine % pour ceux qui sont \`a l'est
%% \URRennes  % pour ceux qui sont \`a l'ouest
%% \URRhoneAlpes % pour ceux qui sont dans les montagnes
%% \URRocq % pour ceux qui sont au centre de la France
%% \URFuturs % pour ceux qui sont dans le virtuel
%% \URSophia % pour ceux qui sont au Sud.
%%
%% \RCBordeaux % centre de recherche Bordeaux - Sud Ouest
%% \RCLille % centre de recherche Lille Nord Europe
%% \RCParis % Paris Rocquencourt
\RCSaclay % Saclay \^Ile de France
%% \RCGrenoble % Grenoble - Rh\^one-Alpes
%% \RCNancy % Nancy - Grand Est
%% \RCRennes % Rennes - Bretagne Atlantique
%% \RCSophia % Sophia Antipolis M\'editerran\'ee


% \maketitle

\newtheorem{definition}{Definition}[section]
\newtheorem{example}[definition]{Example}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{theorem}[definition]{Theorem}


\begin{document}
\sloppy\hbadness=9999
%%
\makeRR   % cas d'un rapport de recherche
% \makeRT % cas d'un rapport technique.
%% a partir d'ici, chacun fait comme il le souhaite


\tableofcontents
\clearpage

\listoffigures

\clearpage

\section{Introduction}

The term \emph{Deductive Verification} denotes the methodology for verifying
that a computer programs is conforming to a specification of its intended
behaviour, with the use of techniques based of automated theorem proving. In
this context, the specification is written in some formal mathematical language,
and the conformity of the code with respect to the specification is
automatically reduced into a set of mathematical formulas than must proved
valid. When the computer programs considered are expressed in programming
language making use of pointers to access their memory heap, the interpretation
of programs into mathematical formulas must rely on a logical description of
such pointers and memory heap: the so-called \emph{memory model}.

Such a memory model cannot be described uniformly among different programming
languages. On the contrary, a specific memory model must be defined for each
programming languages considered. It is even possible to consider several memory
models for different supported of some programming language. \todo{cite
  Frama-C/WP ?}

When reasoning mathematically on pointer programs, the suitability of the memory
model is crucial. One major challenge is the potential \emph{aliasing} of memory
contents: when using pointers, the same memory cell cannot be denoted by
different ``names'',or ``access path''. Modification of the memory content
through one of these names should be taken into account when accessing the same
memory cell via another name. More generally, blocks of memory cells can be read
of written through access path, and these blocks may overlap. The property of
non-overlapping for memory blocks is called their \emph{separation}. Separation
assumptions play an important role when reasoning about pointer
programs~\cite{bornat00mpc}.

The Décysif project gathers partners involved in the design of several programming language involving different approaches for memory access. They develop tools that each contains different approaches for delaing with separation: the tool Why3 for WhyML programs, the SPARK environment for Ada code, the J3 analyzer for C and C++ code, and the Creusot tool for Rust code.

The purpose of this document is to summarise the different memory models
implemented in these tools.

% Generalities about memory models ?

TODO: Overview of the document.

\section{The memory model of Why3}

The region-based aliasing analysis of Why3~\cite{gondelman16reg}.

Common current issues. E.g. with specification, with proof automation, with
counterexamples.

\section{The memory model of Spark}

Complete absence of pointers at first, then partial support of pointers, aliased
controlled using a borrow checker~\cite{dross20cav,jaloyan20icfem}.

How the translation to Why3 works.

Common current issues. E.g. with specification, with proof automation, with
counterexamples.

\section{The memory model of J3}

Aims at supporting low-level C code, with arbitrary pointer casts. Inspired by
CompCert V2~\cite{leroy12rr}. Not yet supported but future work: static analysis
of separation similar to the old plugin
Jessie~\cite{hubert2008these,hubert07hav}.

How the translation to Why3 works.

\section{The memory model of Creusot}

Relies on Rust's borrow checker to propose a translation to Why3. mutable
borrows are translated to pairs of current value and future final value
(prophecy).\cite{denis23phd,denis22icfem,matsushita22pldi}.

\section{Conclusions and Future Work}

Future work
\begin{itemize}
\item New intermediate language in Why3, continuation-based. On going thesis of
  Paul Patault.
\item Separation analysis for J3.
\item Support for concurrency in Creusot, Support for ghost aliasing, on-going
  thesis of Arnaud Golfouse.
\item Better support for counterexamples, needs to makes memory models more
  concrete, more ``executable-like''
\end{itemize}

\bibliographystyle{plainurl}
%\bibliography{abbrevs,demons,demons2,demons3,team,crossrefs}
\bibliography{generated,extra}

%\clearpage


\end{document}

% Local Variables:
% mode: latex
% TeX-master: t
% TeX-PDF-mode: t
% mode: flyspell
% ispell-local-dictionary: "british"
% fill-column: 80
% End:
